---
title: "coursework_cleaning"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# DATA PREPROCESSING

Installing the required packages
```{r}
# Install the follwing packages if not installed already 
install.packages("tidyverse")
install.packages("dplyr")
install.packages("rmarkdown")
```

importing packages
```{r}
library(rmarkdown)
library(dplyr)

```


importing the datasets and assigning NA and empty values with true NA so that missing values can be removed easily
```{r}
# importing the datasets and replacing NA and empty values with NA
dataset2006 <- read.csv("datasets/2006.csv", na.strings = c("", "NA"))
dataset2007 <- read.csv("datasets/2007.csv", na.strings = c("", "NA"))
```


Checking the class of dataset2006 and dataset2007 
```{r}
class(dataset2006)
class(dataset2007)
```


Finding the dimensions of dataset2006 and dataset2007 to see if they can be merged
```{r}
dim(dataset2006)
dim(dataset2007)
```

Creating the full dataset
```{r}
# Merging the 2006 and 2007 datasets using rbind
full_dataset <- rbind(dataset2006, dataset2007)
```


Finding the dimensions of the full_dataset
```{r}
dim(full_dataset)
```


Removing dataset2006 and dataset2007 to free memory
```{r}
rm(dataset2006)
rm(dataset2007)
```


Finding the column names of the full_dataset
```{r}
colnames(full_dataset)
```


Checking the full_dataset dataframe for null values
```{r}
colSums(is.na(full_dataset))
```


Removing null values from ArrDelay
```{r}
full_dataset <- full_dataset[!is.na(full_dataset$ArrDelay),]
```


Checking for null values
```{r}
colSums(is.na(full_dataset))
```


Dropping the CancellationCode column 
```{r}
full_dataset <- subset(full_dataset, select = -c(CancellationCode))
```


Checking the new shape of full_dataset data frame
```{r}
dim(full_dataset)
```


Viewing the top rows in full_dataset
```{r}
head(full_dataset)
```


Checking for duplicate rows
```{r}
head(full_dataset[duplicated(full_dataset),])
```



Dropping duplicate rows in full_dataset
```{r}
full_dataset <- distinct(full_dataset)
```


Saving the cleaned dataset
```{r}
write.csv(full_dataset,"full_dataset.csv")
```




