---
title: "QUESTION_1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# QUESTION 1

# PREPARING THE DATASET

Installing the required packages
```{r}
# Install the follwing packages if not installed already
install.packages("installr")
install.packages("Rtools")
install.packages("rmarkdown")
install.packages("tidyverse")
install.packages("dplyr")
install.packages("plotly")
install.packages("ggplot2")

```

Importing packages
```{r}
library(rmarkdown)
library(dplyr)
library(plotly)
library(ggplot2)
library(tidyverse)
library(tidyr)
```

Importing the main dataset
```{r}
full_dataset <- read.csv("datasets/full_dataset.csv")
```

Checking the full_dataset
```{r}
# Viewing the first 5 rows of the full_dataset
head(full_dataset)
```

Removing the X column in the full_dataset as it adds no value to our analysis
```{r}
full_dataset <- subset(full_dataset, select = -c(X))

```

Checking available columns
```{r}
colnames(full_dataset)
```


Finding the correlation of ArrDelay and DepDelay
```{r}
# Subsetting only the ArrDelay and DepDelay column and finding the pearson correlation matrix for ArrDelay and DepDelay
cor(subset(full_dataset, select = c(ArrDelay, DepDelay)),method = "pearson")
```

# BEST TIME OF YEAR TO FLY TO MINIMIZE DELAYS

Checking for errors in Month column
```{r}
# Checking for smallest and largest values as well as outliers in Months column

# Obtaining the minimum value in the month column
paste("The smallest value for Months is", min(full_dataset$Month))
# Obtaining the maximum value in the month column
paste("The largest value for Months is", max(full_dataset$Month))
# Obtaining a table for the value counts of each month
table(full_dataset$Month)
```
Making a dataframe for mean and median Arrival Delay and Departure Delay for each month
```{r}
# Using the dplyr pipe operator we first group the data by month and then create summarizing variables which show the mean and median for ArrDelay and DepDelay
full_dataset %>% group_by(Month) %>% 
  summarise(mean_ArrDelay = mean(ArrDelay),
            median_ArrDelay = median(ArrDelay),
            mean_DepDelay = mean(DepDelay),
            median_DepDelay = median(DepDelay), .groups = 'drop')

```
## CONSTRUCTING A LINE PLOT TO SHOW HOW AVERAGE ARRIVAL DELAY VARIES ACROSS MONTHS

Making a variable which holds the dataframe containing the mean Arrival Delay (ArrDelay) grouped by Month
```{r}
# Grouping the data by month and calculating the average ArrDelay for each month and storing it in a variable called full_dataset_meandelay
full_dataset_meandelay <- full_dataset %>% group_by(Month) %>% summarise(mean_ArrDelay_month = mean(ArrDelay))
full_dataset_meandelay
```


Assigning a variable to hold the mean of all Arrival Delays
```{r}
avg_ArrDelay <- mean(full_dataset$ArrDelay)
```


Plotting the line graph
```{r}
# Initializing ggplot with the full_dataset_meandelay data frame and selecting the x axis to be months and y axis to be the mean arrival delay
ggplot(full_dataset_meandelay, aes(x = Month, y = mean_ArrDelay_month)) +
  
  geom_line(colour = "blue") +                                 # Plotting a line graph 
  
  geom_hline(yintercept = avg_ArrDelay,  colour = "red", ) +   # Plotting a reference line (the overall average arrival delay)
  
  scale_x_continuous(breaks = seq(0,12,1)) +                   # Formatting x axis tick labels
  scale_y_continuous(breaks = seq(6,14,1)) +                   # Formatting y axis tick labels
  
  ggtitle("Line Plot for Average Arrival Delay by Month") +    # Giving the graph title
  
  xlab("Months") +                                             # Creating the x-axis labels
  
  ylab("Average Arrival Delay
(Time in minutes)") +                                          # Creating the y-axis labels
  
  # Formatting the titles, texts, background and gridlines
  theme(plot.title=element_text(family="mono", 
                                face="bold", 
                                size=10,
                                hjust = 0.5,
                                margin = margin(t = 0, r = 0, b = 5, l = 0)),
  
        axis.title.x=element_text(family="mono",
                                  size=12,
                                  margin = margin(t = 20, r = 0, b = 10, l = 0)),
  
        axis.title.y=element_text(family="mono",
                                  size=12,
                                  margin = margin(t = 0, r = 20, b = 0, l = 10)),
  
        axis.text.x = element_text(colour="#ef4343",
                                   size = 10),
    
        axis.text.y = element_text(colour="#ffc000",
                                   size = 10),
  
        panel.background = element_rect(fill = "white", 
                                        colour = "black",
                                        size = 1, 
                                        linetype = "solid"),
  
        panel.grid.major.y = element_line(size = 0.25, 
                                          linetype = 'dotted',
                                          colour = "grey"), 
  
        panel.grid.minor.y = element_blank(),
  
        panel.grid.major.x = element_line(size = 0.25, 
                                          linetype = 'dotted',
                                          colour = "grey"))
```
Saving the line graph
```{r}
ggsave('output/line_plot_for_average_arrdelay_vs_month.png')
```
Removing full_dataset_meandelay from memory
```{r}
rm(full_dataset_meandelay)
```




## CONSTRUCTING A LINE PLOT FOR THE CAUSAL FACTORS OF AVERAGE DELAY BY MONTH

Filtering the dataset to obtain relevant delays
```{r}
delay_month_df <- subset(full_dataset, select = c(Month, CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay))
```


Making a data frame which holds the average CarrierDelay, WeatherDelay, NASDelay, SecurityDelay and LateAircraftDelay grouped by month
```{r}
delay_month_df <- delay_month_df %>% group_by(Month) %>% summarise(mean_CarrierDelay = mean(CarrierDelay),
                                                                   mean_WeatherDelay = mean(WeatherDelay),
                                                                   mean_NASDelay = mean(NASDelay),
                                                                   mean_SecurityDelay = mean(SecurityDelay),
                                                                   mean_LateAircraftDelay = mean(LateAircraftDelay)) 
delay_month_df
```


Converting delay_month_df to long format so that each delay can be categorized and visualized by colour using ggplot
```{r}
# Making a variable df2 which holds the long format of the delay_month_df variable
df2 <- delay_month_df %>% pivot_longer(cols = c(mean_CarrierDelay, mean_WeatherDelay, mean_NASDelay, mean_SecurityDelay, mean_LateAircraftDelay))
```

Checking the type and column names of df2
```{r}
typeof(df2)
colnames(df2)
```


Plotting the line graphs 
```{r}
# Initializing ggplot with the df2 data frame and selecting the x axis to be months and y axis to be the value of the delay and categorizing each type of delay by colour
ggplot(df2, aes(x = Month, y = value, colour = name)) +
  
  geom_line(size=0.75)+                                                                # Plotting a line graph
  
  scale_x_continuous(breaks = seq(0,12,1)) +                                           # Formatting x axis tick labels
  scale_y_continuous(breaks = seq(0,7,1)) +                                            # Formatting y axis tick labels
  
  ggtitle("Line Plot for Causal Factors of Average Arrival Delay by Month") +          # Setting the graph title
  
  xlab("Months") +                                                                     # Creating the x-axis labels
  
  ylab("Average Arrival Delay
(Time in minutes)") +                                                                  # Creating the y-axis labels
  
  # Formatting the titles, texts, background and gridlines
  theme(plot.title=element_text(family="mono", 
                                face="bold", 
                                size=10,
                                hjust = 0.5,
                                margin = margin(t = 0, r = 0, b = 5, l = 0)),
        
        axis.title.x=element_text(family="mono",
                                  size=12,
                                  margin = margin(t = 20, r = 0, b = 10, l = 0)),
        
        axis.title.y=element_text(family="mono",
                                  size=12,
                                  margin = margin(t = 0, r = 20, b = 0, l = 10)),
        
        axis.text.x = element_text(colour="#ef4343",
                                   size = 10),
        
        axis.text.y = element_text(colour="#ffc000",
                                   size = 10),
        
        panel.background = element_rect(fill = "white", 
                                        colour = "black",
                                        size = 1, 
                                        linetype = "solid"),
        
        panel.grid.major.y = element_line(size = 0.25, 
                                          linetype = 'dotted',
                                          colour = "grey"), 
        
        panel.grid.minor.y = element_blank(),
        
        panel.grid.major.x = element_line(size = 0.25, 
                                          linetype = 'dotted', colour = "grey"))
```


Saving the line graph
```{r}
ggsave('output/line_plot_for_cause_of_average_arrdelay_vs_month.png')
```


Removing unnecessary variables from memory
```{r}
rm(df2)
rm(delay_month_df)
```



# BEST TIME OF WEEK TO FLY TO MINIMIZE DELAYS

Checking for errors in DayOfWeek column
```{r}
# Checking for smallest and largest values aswell as outliers in DayOfWeek column

# Obtaining the minimum value in the DayOfWeek column
paste("The smallest value for DayOfWeek is", min(full_dataset$DayOfWeek))
# Obtaining the maximum value in the DayOfWeek column
paste("The largest value for DayOfWeek is", max(full_dataset$DayOfWeek))
# Obtaining a table for the value counts of each Day Of the Week
table(full_dataset$DayOfWeek) 
```
Constructing a dataframe for mean and median Arrival Delay and Departure Delay which are grouped by DayOfWeek
```{r}
full_dataset %>% group_by(DayOfWeek) %>% 
  summarise(mean_ArrDelay = mean(ArrDelay),
            median_ArrDelay = median(ArrDelay),
            mean_DepDelay = mean(DepDelay),
            median_DepDelay = median(DepDelay), .groups = 'drop') 
```
## CONSTRUCTING A LINE PLOT TO SHOW HOW AVERAGE ARRIVAL DELAY VARIES ACROSS DAYS OF THE WEEK

Making a variable which holds the dataframe containing the mean Arrival Delay (ArrDelay) grouped by DayOfWeek
```{r}
full_dataset_meandelay <- full_dataset %>% group_by(DayOfWeek) %>% summarise(mean_ArrDelay_DayOfWeek = mean(ArrDelay))
full_dataset_meandelay

```
Plotting the line graph
```{r}
# Initializing ggplot with the full_dataset_meandelay data frame and selecting the x axis to be days of the week and y axis to be the mean arrival delay by day of the week
ggplot(full_dataset_meandelay, aes(x = DayOfWeek, y = mean_ArrDelay_DayOfWeek)) +
  
  geom_line(colour = "blue") +                                                    # Plotting the line graph
  
  geom_hline(yintercept = avg_ArrDelay,  colour = "red", ) +          # Plotting a reference line (overall average arrival delay)
  
  scale_x_continuous(breaks = seq(0,12,1)) +
  scale_y_continuous(breaks = seq(6,14,1)) +
  
  ggtitle("Line Plot for Average Arrival Delay by Day of the Week") +             # Setting the title
  
  xlab("Day of the Week") +                                                       # Creating the x-axis labels
  
  ylab("Average Arrival Delay
(Time in minutes)") +                                                             # Creating the y-axis labels
  
  
  # Formatting the titles, texts, background and gridlines
  theme(plot.title=element_text(family="mono", 
                                face="bold", 
                                size=10,
                                hjust = 0.5,
                                margin = margin(t = 0, r = 0, b = 5, l = 0)),
        
        axis.title.x=element_text(family="mono",
                                  size=12,
                                  margin = margin(t = 20, r = 0, b = 10, l = 0)),
        
        axis.title.y=element_text(family="mono",
                                  size=12,
                                  margin = margin(t = 0, r = 20, b = 0, l = 10)),
        
        axis.text.x = element_text(colour="#ef4343",
                                   size = 10),
        
        axis.text.y = element_text(colour="#ffc000",
                                   size = 10),
        
        panel.background = element_rect(fill = "white", 
                                        colour = "black",
                                        size = 1, 
                                        linetype = "solid"),
        
        panel.grid.major.y = element_line(size = 0.25, 
                                          linetype = 'dotted',
                                          colour = "grey"), 
        
        panel.grid.minor.y = element_blank(),
        
        panel.grid.major.x = element_line(size = 0.25, 
                                          linetype = 'dotted',
                                          colour = "grey"))  
```


Saving the line graph
```{r}
ggsave('output/line_plot_for_average_arrdelay_vs_dayofweek.png')
```


removing full_dataset_meandelay from memory
```{r}
rm(full_dataset_meandelay)
```




## CONSTRUCTING A LINE PLOT FOR THE CAUSAL FACTORS OF AVERAGE DELAY BY DAY OF THE WEEK

Filtering the dataset to obtain relevant delays
```{r}
delay_DayOfWeek_df <- subset(full_dataset, select = c(DayOfWeek, CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay))
```


Making a dataframe which holds the average CarrierDelay, WeatherDelay, NASDelay, SecurityDelay and LateAircraftDelay grouped by month
```{r}
delay_DayOfWeek_df <- delay_DayOfWeek_df %>% group_by(DayOfWeek) %>% summarise(mean_CarrierDelay = mean(CarrierDelay),
                                                                       mean_WeatherDelay = mean(WeatherDelay),
                                                                       mean_NASDelay = mean(NASDelay),
                                                                       mean_SecurityDelay = mean(SecurityDelay),
                                                                       mean_LateAircraftDelay = mean(LateAircraftDelay)) 
delay_DayOfWeek_df
```



Converting delay_DayOfWeek_df to long format so that each delay can be categorized and visualized by colour using ggplot
```{r}
# Making a variable df2 which holds the long format of the delay_DayOfWeek_df variable
df2 <- delay_DayOfWeek_df %>% pivot_longer(cols = c(mean_CarrierDelay, mean_WeatherDelay, mean_NASDelay, mean_SecurityDelay, mean_LateAircraftDelay))
```
Checking the type and column names of df2
```{r}
typeof(df2)
colnames(df2)
```


Plotting the line graphs 
```{r}
# Initializing ggplot with the df2 data frame and selecting the x axis to be DayOfWeek and y axis to be the value of the delay and categorizing each type of delay by colour
ggplot(df2, aes(x = DayOfWeek, y = value, colour = name)) +
  
  geom_line(size=0.75)+                                                                 # Plotting the line graphs
  
  scale_x_continuous(breaks = seq(0,7,1)) +
  scale_y_continuous(breaks = seq(0,7,1)) +                                             # Formatting the x and y axis tick labels
  
  ggtitle("Line Plot for Causal Factors of Average Arrival Delay by Day of the Week") + # Setting the title
  
  xlab("Day of the Week") +                                                             # Creating the x-axis labels
  
  ylab("Average Arrival Delay
(Time in minutes)") +                                                                   # Creating the y-axis labels
  
  # Formatting the titles, texts, background and gridlines
  theme(plot.title=element_text(family="mono", 
                                face="bold", 
                                size=10,
                                hjust = 0.5,
                                margin = margin(t = 0, r = 0, b = 5, l = 0)),
        
        axis.title.x=element_text(family="mono",
                                  size=12,
                                  margin = margin(t = 20, r = 0, b = 10, l = 0)),
        
        axis.title.y=element_text(family="mono",
                                  size=12,
                                  margin = margin(t = 0, r = 20, b = 0, l = 10)),
        
        axis.text.x = element_text(colour="#ef4343",
                                   size = 10),
        
        axis.text.y = element_text(colour="#ffc000",
                                   size = 10),
        
        panel.background = element_rect(fill = "white", 
                                        colour = "black",
                                        size = 1, 
                                        linetype = "solid"),
        
        panel.grid.major.y = element_line(size = 0.25, 
                                          linetype = 'dotted',
                                          colour = "grey"), 
        
        panel.grid.minor.y = element_blank(),
        
        panel.grid.major.x = element_line(size = 0.25, 
                                          linetype = 'dotted', colour = "grey"))
```

Saving the line graph
```{r}
ggsave('output/line_plot_for_cause_of_average_arrdelay_vs_dayofweek.png')
```

Removing unnecessary variables from memory
```{r}
rm(df2)
rm(delay_DayOfWeek_df)
```




# BEST TIME OF DAY TO FLY TO MINIMIZE DELAYS

We analyze using CRSDepTime since customers can only choose from scheduled departure times

Checking for errors in CRSDepTime column

```{r}
# Finding the smallest and largest values CRSDepTime column
paste("The smallest value for CRSDepTime is", min(full_dataset$CRSDepTime))
paste("The largest value for CRSDepTime is", max(full_dataset$CRSDepTime))
```

Checking the type of CRSDepTime
```{r}
typeof(full_dataset$CRSDepTime)
```
 
Making 8 bins which are spaced 3 hours apart to categorize each time of day
```{r}
# Making a list which holds 8 intervals of different times of day of their corresponding Arrival Delays
delay_TimeOfDay = list(full_dataset[(full_dataset$CRSDepTime >= 0000.0 & full_dataset$CRSDepTime < 0300.0),]$ArrDelay, 
                       full_dataset[(full_dataset$CRSDepTime >= 0300.0 & full_dataset$CRSDepTime < 0600.0),]$ArrDelay, 
                       full_dataset[(full_dataset$CRSDepTime >= 0600.0 & full_dataset$CRSDepTime < 0900.0),]$ArrDelay,
                       full_dataset[(full_dataset$CRSDepTime >= 0900.0 & full_dataset$CRSDepTime < 1200.0),]$ArrDelay,
                       full_dataset[(full_dataset$CRSDepTime >= 1200.0 & full_dataset$CRSDepTime < 1500.0),]$ArrDelay,
                       full_dataset[(full_dataset$CRSDepTime >= 1500.0 & full_dataset$CRSDepTime < 1800.0),]$ArrDelay,
                       full_dataset[(full_dataset$CRSDepTime >= 1800.0 & full_dataset$CRSDepTime < 2100.0),]$ArrDelay,
                       full_dataset[(full_dataset$CRSDepTime >= 1800.0 & full_dataset$CRSDepTime <= 2400.0),]$ArrDelay)
```

## CONSTRUCTING A BOXPLOT FOR ARRIVAL DELAY BY TIME OF DAY
```{r}
# Setting the plot to default settings
par(mar = c(6.1, 4.1, 4.1, 4.1), # changing the margins
    lwd = 1, # setting the thickness of the lines
    cex.axis = 0.5 # changing axis label sizes
)

# Plotting the boxplot
box_plot <- boxplot(delay_TimeOfDay, xaxt = "n", ylim=c(-75,75))

# Creating the x axis
axis(side = 1, labels = FALSE)

text(x = 1:length(delay_TimeOfDay),
     y = par("usr")[3] -3.0,                  # Drawing labels at the bottom of the chart.
     labels = c('0000h-0300h',      # Using the names of the 8 time intervals from the delay_TimeOfDay list as the x axis labels.
                '0300h-0600h', 
                '0600h-0900h',
                '0900h-1200h',
                '1200h-1500h',
                '1500h-1800h',
                '1800h-2100h',
                '2100h-0000h'),
     xpd = NA,                                # Changing the clipping region to correct label positioning.
     srt = 90,                                # Rotating the labels by 90 degrees.
     adj = 0.9,                               # Adjusting the labels to 90% right-justified.
     cex = 0.8)                               # Decreasing the label size.
```


Getting the summary statistics from the box plot
```{r}
box_plot$stats
# The row [3,] in the output shows the median of each box plot
```

removing delay_TimeOfDay and box_plot from memory
```{r}
rm(delay_TimeOfDay)
rm(box_plot)
```





































































































