---
title: "QUESTION_2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# QUESTION 2

# PREPARING THE DATASET

Installing the required packages
```{r}
# Install the follwing packages if not installed already
install.packages("installr")
install.packages("Rtools")
install.packages("rmarkdown")
install.packages("tidyverse")
install.packages("dplyr")
install.packages("plotly")
install.packages("ggplot2")
```

importing packages
```{r}
library(dplyr)
library(plotly)
library(ggplot2)
```


importing the main dataset and assigning NA and empty values with true NA so that missing values can be removed easily
```{r}
full_dataset <- read.csv("datasets/full_dataset.csv", na.strings = c("", "NA"))
```

Removing the X column in the full_dataset as it adds no value to our analysis
```{r}
full_dataset <- subset(full_dataset, select = -c(X))

```

Checking available columns in full_dataset
```{r}
colnames(full_dataset)
```

importing the plane dataset and replacing NA and empty values with true NA so that missing values can be removed easily
```{r}
plane_dataset <- read.csv("datasets/plane-data.csv", na.strings = c("", "NA"))
```

Checking available columns in plane_dataset
```{r}
colnames(plane_dataset)
```


Removing unnecessary columns in the full_dataset
```{r}
# Selecting only the TailNum, ArrDelay, DepDelay columns in the full_dataset
full_dataset <- subset(full_dataset, select = c(TailNum, ArrDelay, DepDelay))
head(full_dataset)
```


Removing unnecessary columns in the plane_dataset
```{r}
# Selecting only the tailnum and year columns in the plane_dataset
plane_dataset <- subset(plane_dataset, select = c(tailnum, year))
head(plane_dataset)
```


Checking for null values in plane_dataset

```{r}
colSums(is.na(plane_dataset))
```

Removing the null values in the plane_dataset
```{r}
plane_dataset <- plane_dataset[!is.na(plane_dataset$year),]
colSums(is.na(plane_dataset))

```

Making a copy of the plane dataset is not required in R but it is required in python.

Renaming the plane_dataset tail number column to match that of the full_dataset tail number column so that merging is possible.
```{r}
# Renaming the 'tailnum' column to 'TailNum' in the plane_dataset
plane_dataset_1 <- plane_dataset %>% rename(TailNum=tailnum)
head(plane_dataset_1)
```

Removing unnecessary variables
```{r}
rm(plane_dataset)
```


Merging plane_dataset_1 with the full_dataset based on Tail Number
```{r}
merged_dataset <- merge(full_dataset, plane_dataset_1, by='TailNum')
head(merged_dataset)
```


Checking for null values in the merged_dataset
```{r}
colSums(is.na(merged_dataset))
```


Checking the dtypes of the merged_dataset
```{r}
str(merged_dataset)
```


Since the year is an object we check the year column for non-numerical values and unfeasible year values
```{r}
table(merged_dataset$year)
```


Removing the '0000' and 'None' values from the year column in the merged_dataset
```{r}
merged_dataset <- merged_dataset[!(merged_dataset$year == '0000' | merged_dataset$year == 'None' ),]
table(merged_dataset$year)
```

Now the year column can be converted to an integer as it is now only comprised of numerics 
```{r}
# Converting the year variable from an object to an integer
merged_dataset$year <- as.integer(merged_dataset$year)
# Checking the dtypes of the merged_dataset to make sure year is now an integer variable
str(merged_dataset)
```



# CONSTRUCTING A LINE PLOT TO COMPARE THE AVERAGE ARRIVAL DELAY AGAINST PLANE MANUFACTURE YEAR

Making a variable which holds the data frame containing the mean Arrival Delay (ArrDelay) grouped by Plane Manufactured Year (year)
```{r}
avg_ArrDelay_plane_year <- merged_dataset %>% group_by(year) %>% summarise(mean_ArrDelay = mean(ArrDelay), .groups = 'drop')
```


Assigning a variable to hold the overall mean of Arrival Delays
```{r}
avg_ArrDelay <- mean(merged_dataset$ArrDelay)
```

Plotting the line graph
```{r}
# Initializing ggplot with the avg_ArrDelay_plane_year data frame and selecting the x axis to be Plane Manufacture Years (year) and y axis to be the mean arrival delay
ggplot(avg_ArrDelay_plane_year, aes(x = year, y = mean_ArrDelay)) +
  
  geom_line(colour = "blue") +                                    # Plotting a line graph 
  
  geom_hline(yintercept = avg_ArrDelay,  colour = "red", ) +      # Plotting a reference line (the overall average arrival delay)
  
  scale_x_continuous(breaks = seq(1955,2011,5)) +                 # Formatting x axis tick labels
  scale_y_continuous(breaks = seq(6,14,1)) +                      # Formatting y axis tick labels
  
  ggtitle("Line Plot for Average Arrival Delay by Plane Manufactured Year") +             # Giving the graph a title
  
  xlab("Plane Manufactured Year") +                               # Setting the x-axis labels
  
  ylab("Average Arrival Delay
(Time in minutes)") +                                             # Setting the y-axis labels
  
  # Formatting the titles, texts, background and gridlines
  theme(plot.title=element_text(family="mono", 
                                face="bold", 
                                size=10,
                                hjust = 0.5,
                                margin = margin(t = 0, r = 0, b = 5, l = 0)),
        
        axis.title.x=element_text(family="mono",
                                  size=12,
                                  margin = margin(t = 20, r = 0, b = 10, l = 0)),
        
        axis.title.y=element_text(family="mono",
                                  size=12,
                                  margin = margin(t = 0, r = 20, b = 0, l = 10)),
        
        axis.text.x = element_text(colour="#ef4343",
                                   size = 10),
        
        axis.text.y = element_text(colour="#ffc000",
                                   size = 10),
        
        panel.background = element_rect(fill = "white", 
                                        colour = "black",
                                        size = 1, 
                                        linetype = "solid"),
        
        panel.grid.major.y = element_line(size = 0.25, 
                                          linetype = 'dotted',
                                          colour = "grey"), 
        
        panel.grid.minor.y = element_blank(),
        
        panel.grid.major.x = element_line(size = 0.25, 
                                          linetype = 'dotted',
                                          colour = "grey"))
```
  

Saving the line graph
```{r}
ggsave('output/line_plot_for_average_arrdelay_vs_planemanufactureyear.png')
```


Removing unnecessary variables from memory
```{r}
rm(avg_ArrDelay_plane_year)
rm(merged_dataset)
rm(plane_dataset_1)
rm(full_dataset)
rm(avg_ArrDelay)
```





















